---
title: "Introduction to corncob"
author: "Bryan D Martin"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Introduction to corncob}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This tutorial is intended for STAMPS 2019.

When studying dysbiosis, the focus is often on identifying differentially abundant taxa. However, microbial imbalance can also affect the stability of taxa. Here, we introduce \texttt{corncob}, an individual taxon regression model that uses abundance tables and sample data. \texttt{corncob} is able to model differential abundance and differential variability. \texttt{corncob} is appropriate for both rare and common taxa, excessive zeros, overdispersion, and inconsistent sequencing depths.

To begin, we load an example data set as a \texttt{phyloseq} object.

```{r, message = FALSE}
library(corncob)
library(phyloseq)
data(soil_phylo)
```

If you are unfamiliar with \texttt{phyloseq}, we can view a description of the data using:

```{r}
soil_phylo
```

Let's examine a small subset of our data in more detail.
```{r}
otu_table(soil_phylo)[1:5, 1:5]
sample_data(soil_phylo)[1:5, ]
```
